%
%

\paragraph{Rigid grammars}
We first define the restricted notion of grammars that we study.
%
Consider an annotated grammar $\cG = (N, \Sigma, \Omega, P, S)$, and a string
$\gamma\in (\Sigma \cup (\Sigma \times \Omega)\cup N)^*$ of nonterminals and
of terminals which may carry an annotation in~$\Omega$. We will be interested in
the \emph{shape}
of~$\gamma$, written $\shape(\gamma)$: it is the string over $\{0, 1\}$
obtained by replacing every nonterminal of~$N$ in~$\gamma$ by~$1$ and replacing
all terminals (annotated or not) by~$0$: note that $|\shape(\gamma)| = |\gamma|$.

We then say that an annotated grammar $\cG$ is \emph{rigid}
%
%
	%
        if for every string
        $w \in \Sigma^*$, all derivations from the start symbol~$S$ of~$\cG$ to
        an annotated string $\hat{w}$ of~$w$ have the same sequence of shapes
        (remember that we only consider leftmost derivations).
        Formally, there exists a sequence $s_1, \ldots, s_k \in \{0,
        1\}^*$ depending only on~$w$ such that for any derivation 
        $S = \alpha_1 \der{\cG} \alpha_2 \der{\cG} \ldots \der{\cG} \alpha_m =
        \hat{w}$ with $\str(\hat{w}) = w$, we have $m = k$ and $\mathrm{shape}(\alpha_i) = s_i$ for all $1
        \leq i \leq k$.
	%
	%
	%
%

Intuitively, the sequence of shapes of a derivation describes the
skeleton of the corresponding derivation tree. Thus, a rigid annotated
grammar is one where, for each unannotated string, all derivation trees for all
annotations of the string are isomorphic (ignoring the labels of 
nonterminals and the annotation of terminals).

Now we restate Proposition~\ref{gram:prp:2nf} while including the angle of rigid grammars.

\begin{proposition}[Proposition~\ref{gram:prp:2nf}]
  \label{gram:prp:2nfrig}
  Given any annotated grammar\/ $\mathcal{G}$, we can compute in linear time an annotated
  grammar $\mathcal{G}'$ in 2NF such that $\mathcal{G}$ and $\mathcal{G}'$ are
  equivalent. Furthermore, if $\mathcal{G}$ is unambiguous (resp. rigid) then $\mathcal{G}'$ is unambiguous (resp. rigid) as well.
\end{proposition}
\begin{proof}
  \input{./proofs/grammars/arity2.tex}
\end{proof}









\paragraph{Rigidity vs unambiguity}
Unambiguity and rigidity for annotated grammars seem
incomparable: unambiguity imposes that every annotation is produced by only one
derivation, whereas rigidity imposes that all derivations across all annotations
have the same shape (but the same annotation may be obtained multiple times).

However, it turns out that, on rigid grammars, we can impose unambiguity without
loss of generality: all rigid grammars can be converted to equivalent rigid and
unambiguous grammars.

\begin{theorem}
  \label{gram:thm:profileu-iou} For any rigid grammar
	$\cG$ we can build an equivalent rigid and
	unambiguous grammar $\cG'$. The transformation runs in exponential time,
        i.e., time $\cO(2^{|\cG|^c})$ for some $c> 0$.
\end{theorem}
\begin{proof}
  \input{./proofs/grammars/quadres1}
\end{proof}

% The transformation to impose unambiguity goes via a notion of annotated pushdown
% automata (introduced in the next section), and is inspired by the
% determinization procedure for visibly pushdown automata~\cite{alur2004visibly}, even though rigid
% grammars generally do not define visibly pushdown languages. The transformation
% comes at a cost, as it will generally blow up the size of the grammar
% exponentially.



\paragraph{Expressiveness of rigid grammars}
Armed with Theorem~\ref{gram:thm:profileu-iou}, we study what is the expressive power
of rigid grammars. For this, let us first go back to the setting without
annotations. Theorem~\ref{gram:thm:profileu-iou} tells us that for (unannotated) CFGs the rigidity
requirement is equivalent to the usual unambiguity requirement: each accepted
word has a unique derivation.
Now, for the case of an annotated grammar $\cG$,
rigidity additionally imposes the requirement
that all annotations of an input string have the same parse tree.
In particular, the language of the strings where $\cG$ accepts some
annotation must be recognizable by a rigid (unannotated) CFG, hence an unambiguous CFG (by
Theorem~\ref{gram:thm:profileu-iou}). Formally:

\begin{proposition}
  \label{gram:prp:baseunambig}
   For a rigid grammar $\cG$, let $L'$ be the set of strings with nonempty output, i.e., $L' = \{w \mid \sem{\cG}(w) \not = \emptyset\} $. Then $L'$ is
  recognized by an unambiguous CFG.
\end{proposition}
\begin{proof}
  \input{./proofs/grammars/quadres2.tex}
\end{proof}

This yields concrete examples of languages (on the empty annotation alphabet) that cannot be recognized by a
rigid annotated grammar, e.g., inherently ambiguous context-free languages
such as $L_\mathrm{a} = \{a^i b^j c^k \mid i, j, k \geq 1 \land (i = j \lor j =
k)\}$ on~$\{a, b, c\}^*$~\cite{maurer1969direct}.
%
%
%
%
Proposition~\ref{gram:prp:baseunambig} also implies that we cannot decide if the language
of an annotated grammar can be expressed instead by a rigid grammar,
or if an annotated grammar is rigid:

\begin{proposition}
  \label{gram:prp:unambundec2}
  Given an unannotated grammar $\cG$, it is undecidable to determine whether $\cG$
  is rigid, and it is undecidable to determine whether there
  is some equivalent rigid grammar $\cG'$.
\end{proposition}
\begin{proof}
  \input{./proofs/grammars/quadres3.tex}
\end{proof}

These undecidability results make 
rigid grammars less appealing, but note that our enumeration algorithm for such
grammars applies in particular to decidable grammar classes which are
designed to ensure rigidity. For instance, this would be the case
of grammars arising from visibly pushdown automata, which we discuss in more
detail in the next section.

%
%
%
%
%
%
%
%
%
%
%
%
%
%
%

